### 2022-07-25 paper/localpsf.tex
edits from discussion with Tucker

By ``matrix free,'' we mean that we have a black-box computational procedure through which we may apply $\Aop$ and its transpose\footnote{Recall that the transpose $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(v) = \left(\Aop v\right)(u)$ for all $u,v \in L^2(\Omega)$, where the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(v)$ is the result of applying that linear functional to $v \in L^2(\Omega)$, and similar for the analogous operations with $\Aop^T$. }, $\Aop^T$, to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv,
\end{equation} 
but we do not have any further access to $\Aop$. In particular, we cannot easily access entries of $\Aop$'s integral kernel. In applications, evaluating the maps in \eqref{eq:eval_maps} may involve iteratively solving a large linear system, timestepping, or performing some other costly computational procedure.

@@

The central challenge we face is that existing scalable high-rank approximation methods typically require the ability to rapidly evaluate arbitrary entries of $\Aop$'s integral kernel, and we cannot do this because $\Aop$ is only accessible matrix-free. We address this challenge using \emph{impulse response interpolation}. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(v) = \left(\rho^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate translated versions of these impulse responses to approximate arbitrary entries of the operator's integral kernel. 
%This allows fast evaluation of approximate matrix entries, which are needed by high-rank approximation methods, in particular, classical H-matrix construction methods.

@@

Let $\Omega \subset \mathbb{R}^\gdim$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$). We seek to approximate integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ that may be written in the following form:
\begin{equation}
\label{eq:kernel_representation}
(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy.
\end{equation}
Here the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(v)$ is the result of applying that linear functional to $v \in L^2(\Omega)$.
The function $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel, which exists in principle but is not easily accessible to us. In this section we describe how to extend the action of $\Aop$ to distributions, which allows us to define the impulse responses (Section \ref{sec:impulse_response_2}). Then we state the conditions on $\Aop$ that our method requires (Section \ref{sec:conditions_2}). Finally, we detail finite element discretization (Section \ref{sec:finite_element_kernel}).

@@

The action of $\Aop$ may be extended to distributions if the kernel $\Aker$ is sufficiently regular. We derive the extension as follows. Suppose $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(v) &= \int_\Omega \int_\Omega v(y) \Aker(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega v(y) \int_\Omega \Aker(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega v(y) \genericdistribution\left(\Aker(y,\cdot)\right) dy, \label{eq:action_on_distribution}
	\end{align}
\end{subequations}
where $\Aker(y,\cdot)$ denotes the function $x \mapsto \Aker(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions, and instead suppose $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ is a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in \eqref{eq:extension_to_distributions} is not valid. However, if $\Aker$ is sufficiently regular such that the function $y \mapsto \rho\left(\Aker(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the right hand side of \eqref{eq:action_on_distribution} is well-defined. Hence, we \emph{define} the action of $\Aop$ on the distribution $\rho$ to be the right hand side of \eqref{eq:action_on_distribution}. For convenience of notation, we denote this action by ``$\Aop \genericdistribution^*$,'' even if $\rho^*$ does not exist.

Let $\delta_x$ denote the delta distribution\footnote{Recall that the delta distribution $\delta_x:\mathcal{D}(\Omega)\rightarrow \mathbb{R}$ is the linear functional defined by $\delta_x(v) = v(x)$ for all $v\in \mathcal{D}(\Omega)$.} (i.e., point source, impulse) centered at the point $x \in \Omega$.
The \emph{impulse response} of $\Aop$ associated with $x$ is the function $\impulseresponse_x:\Omega \rightarrow \mathbb{R}$, 
\begin{equation}
\label{eq:impulse_response_delta_action}
\impulseresponse_x := \left( \Aop \delta_x^* \right)^*,
\end{equation}
that is formed by applying $\Aop$ to $\delta_x$, then taking the Riesz representation of the resulting linear functional. Using \eqref{eq:action_on_distribution} and the definition of the delta distribution, we see that $\phi_x$ may also be written as the function
\begin{equation}
\label{eq:impulse_response_defn}
\impulseresponse_x(y) = \Aker(y, x),
\end{equation} 
which results from keeping $x$ fixed and viewing $\Aker(y,x)$ as a function of $y$.

@@

Let $\febasis_1, \febasis_2, \dots, \febasis_\fedim$ be a set of continuous Galerkin finite element basis functions used to discretize the problem on a mesh with mesh size parameter $h$, let $V_h := \Span\left(\febasis_1, \febasis_2, \dots, \febasis_\fedim\right)$ be the corresponding finite element space under the $L^2$ inner product, and let $p_i \in \mathbb{R}^\gdim$, $i=1,\dots, \fedim$ be the Lagrange nodes associated with the functions $\febasis_i$. We assume that the finite element basis satisfies the Kronecker property, i.e., $\febasis_i(p_i)=1$ and $\febasis_i(p_j)=0$ if $i\neq j$. 
For $u_h \in V_h$ we write $\bm{u} \in \mathbb{R}^\nsamplepts_\massmatrix$ to denote the coefficient vector for $u_h$ with respect to the finite element basis, i.e.,
\begin{equation}
\label{eq:fem_coeff_basis}
u_h(x) = \sum_{i=1}^\fedim \bm{u}_i \febasis_i(x).
\end{equation}
Linear functionals $\rho_h \in V_h'$ have coefficient dual vectors $\boldsymbol{\rho}\in \mathbb{R}^\nsamplepts_{\massmatrix^{-1}}$, with entries 
\begin{equation*}
\boldsymbol{\rho}_i = \rho_h(\febasis_i), \qquad i=1,\dots,\nsamplepts. 
\end{equation*}
Here $\massmatrix \in \mathbb{R}^{\fedim \times \fedim}$ denotes the sparse finite element mass matrix which has entries
\begin{equation*}
\massmatrix_{ij}=\int_\Omega \febasis_i(x) \febasis_j(x) dx, \qquad i,j=1,\dots,\fedim,
\end{equation*}
the space $\mathbb{R}^\fedim_\massmatrix$ is $\mathbb{R}^\fedim$ with the inner product $(\mathbf{u},\mathbf{v})_\massmatrix := \mathbf{u}^T \massmatrix \mathbf{v}$, and $\mathbb{R}^\fedim_{\massmatrix^{-1}}$ is the analogous space with $\massmatrix^{-1}$ replacing $\massmatrix$. Direct calculation shows that $\mathbb{R}^\fedim_\massmatrix$ and $\mathbb{R}^\fedim_{\massmatrix^{-1}}$ are isomorphic to $V_h$ and $V_h'$ as Hilbert spaces, respectively.

After discretization, the operator $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ is replaced by an operator $A_h:V_h \rightarrow V_h'$, which becomes an operator
\begin{equation*}
\mathbf{A}:\mathbb{R}^\fedim_\massmatrix \rightarrow \mathbb{R}^\fedim_{\massmatrix^{-1}}
\end{equation*}
under the isomorphism discussed above. Our method is agnostic to the computational procedure for approximating $\Aop$ with $\mathbf{A}$. What is important is that we do not have direct access to matrix entries $\mathbf{A}_{ij}$. Rather, we have a computational procedure that allows us to compute matrix-vector products $\mathbf{u}\mapsto \mathbf{A}\mathbf{u}$ and $\mathbf{v}\mapsto \mathbf{A}^T\mathbf{v}$, and computing these matrix-vector products is costly.
Of course, matrix entries can be computed via matrix-vector products as $\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i$,
where $\mathbf{e}_j=(0,\dots,0,1,0,\dots,0)^T$ is the length $\fedim$ unit vector with one in the $j$th coordinate and zeros elsewhere. But computing the matrix-vector product $\mathbf{e}_j \mapsto \mathbf{A}\mathbf{e}_j$ is costly, and therefore wasteful if we do not use other matrix entries in the $j$th column of $\mathbf{A}$. Hence, methods for approximating $\mathbf{A}$ are computationally intractable if they require accessing scattered matrix entries from many different rows and columns of $\mathbf{A}$.

The operator $A_h:V_h \rightarrow V_h'$ can be written in integral kernel form as
\begin{equation}
(A_h u_h)(v_h) := \int_\Omega \int_\Omega v_h(y) \Aker_h(y,x) u_h(x) dx dy
\end{equation}
for some kernel $\Aker_h$ which we do not know, and which may differ from $\Aker$ due to discretization error. Since the functions in $V_h$ are continuous at $x$, the delta distribution $\delta_x$ is a continuous linear functional on $V_h$, which has a discrete dual vector $\boldsymbol{\delta}_x \in \mathbb{R}^\fedim_{\massmatrix^{-1}}$ with entries $\left(\boldsymbol{\delta}_x\right)_i = \febasis_i(x)$ for $i=1,\dots,\fedim$. Additionally, it is straightforward to verify that the Riesz representation, $\rho_h^* \in V_h$, of a functional $\rho \in V_h'$ has coefficient vector
\begin{equation*}
\boldsymbol{\rho}^* = \massmatrix^{-1} \boldsymbol{\rho}.
\end{equation*}
Therefore, the formula for the impulse response from \eqref{eq:impulse_response_delta_action} becomes 
\begin{equation}
\label{eq:discrete_kernel}
\boldsymbol{\impulseresponse}_x = \left(A_h \delta_x^*\right)^* =  \massmatrix^{-1}\mathbf{A} \massmatrix^{-1} \boldsymbol{\delta}_x,
\end{equation}
and the $(y,x)$ kernel entry of $\Aker_h$ may be written as
\begin{equation}
\label{eq:approx_kernel_entry_81}
	\Aker_h(y,x) = \boldsymbol{\delta}_y^T \boldsymbol{\impulseresponse}_x = \boldsymbol{\delta}_y^T \massmatrix^{-1}\mathbf{A} \massmatrix^{-1} \boldsymbol{\delta}_x.
\end{equation}
Now define $\mathbf{\Aker} \in \mathbb{R}^{\fedim \times \fedim}$ to be the following dense matrix of kernel entries evaluated at all pairs of Lagrange nodes:
\begin{equation}
\label{eq:Akerpcmat_entries}
\mathbf{\Aker}_{ij} := \Aker_h(p_i, p_j).
\end{equation}
Because of the Kronecker property of the finite element basis, we have $\boldsymbol{\delta}_{p_i} = \mathbf{e}_i$. Thus from \eqref{eq:approx_kernel_entry_81} we have
$
	\Aker_h(p_i,p_j) = \left(\massmatrix^{-1}\mathbf{A} \massmatrix^{-1}\right)_{ij},
$
which implies
\begin{equation}
\label{eq:boldA}
	\mathbf{A} = \massmatrix \mathbf{\Aker} \massmatrix.
\end{equation}
Broadly, our method will construct an H-matrix approximation of $\mathbf{A}$ by forming H-matrix approximations of $\boldsymbol{\Aker}$ and $\massmatrix$ (or a lumped mass version of $\massmatrix$), then multiplying these matrices with H-matrix methods per \eqref{eq:boldA}. The central challenge is that classical H-matrix construction methods require access to arbitrary matrix entries $\mathbf{\Aker}_{ij}$, but these matrix entries are not easily accessible. The bulk of our method is therefore dedicated to forming approximations of these matrix entries that can be evaluated rapidly.

@@

The direct approach to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ is to apply $\Aop$ to a point source centered at $x$ to obtain $\impulseresponse_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\impulseresponse_x$ to determine $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$. However, this direct approach is not feasible because we need to know $V(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ before we compute $\phi_x$, in order choose the point $x$. Computing $\phi_x$ in order to determine $V(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ would be extremely computationally expensive, and defeat the purpose of our algorithm, which is to reduce the computational cost by computing impulse responses in batches. Fortunately, it is possible to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ indirectly, \emph{for all points $x\in\Omega$ simultaneously}, by applying $\Aop^T$ to one constant function, $\gdim$ linear functions, and $\gdim(\gdim+1)/2$ quadratic functions (e.g., 6 total operator applies in two spatial dimensions and 10 in three spatial dimensions). This may be motivated by analogy to matrices. If $\mathbf{B}\in \mathbb{R}^{\fedim \times \fedim}$ is a matrix with $i^\text{th}$ column $\mathbf{b}_i$ and $\mathbf{f} \in \mathbb{R}^\fedim$, then
\begin{equation*}
\mathbf{B}^T \mathbf{f} = \begin{bmatrix}
\horzbar & \mathbf{b}_1^T & \horzbar \\
\horzbar & \mathbf{b}_2^T & \horzbar \\
& \vdots & \\
\horzbar & \mathbf{b}_\fedim^T & \horzbar
\end{bmatrix}
\mathbf{f} = 
\begin{bmatrix}
\mathbf{b}_1^T \mathbf{f} \\
\mathbf{b}_2^T \mathbf{f} \\
\vdots \\
\mathbf{b}_\fedim^T \mathbf{f}
\end{bmatrix}.
\end{equation*}
By computing one matrix-vector product of $\mathbf{B}^T$ with $\mathbf{f}$, we compute the inner product of each column of $\mathbf{B}$ with $\mathbf{f}$ simultaneously. The operator case is analogous, with the impulse response $\phi_x$ taking the place of a matrix column. We have
\begin{equation}
\label{eq:operator_simultaneous_ips}
\left(\Aop^T f\right)^*(x) = \int_\Omega \Aker(y,x) f(y) dy = \left(\phi_x, f\right)_{L^2(\Omega)}.
\end{equation}
By computing one operator application of $\Aop^T$ to $f$, we compute the inner product of each impulse response $\phi_x$ with $f$, for all points $x$ simultaneously.

>>>>

By ``matrix free,'' we mean that we have a black-box computational procedure through which we may apply $\Aop$ and its transpose\footnote{Recall that the transpose $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop w\right)(u)$ for all $u,w \in L^2(\Omega)$, where the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for the analogous operations with $\Aop^T$. }, $\Aop^T$, to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
but we do not have any further access to $\Aop$. In particular, we cannot easily access entries of $\Aop$'s integral kernel. In applications, evaluating the maps in \eqref{eq:eval_maps} may involve iteratively solving a large linear system, timestepping, or performing some other costly computational procedure.

@@

The central challenge we face is that existing scalable high-rank approximation methods typically require the ability to rapidly evaluate arbitrary entries of $\Aop$'s integral kernel, and we cannot do this because $\Aop$ is only accessible matrix-free. We address this challenge using \emph{impulse response interpolation}. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(w) = \left(\rho^*,w\right)_{L^2(\Omega)}$ for all $w \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate translated versions of these impulse responses to approximate arbitrary entries of the operator's integral kernel. 
%This allows fast evaluation of approximate matrix entries, which are needed by high-rank approximation methods, in particular, classical H-matrix construction methods.

@@

Let $\Omega \subset \mathbb{R}^\gdim$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$). We seek to approximate integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ that may be written in the following form:
\begin{equation}
\label{eq:kernel_representation}
(\Aop u)(w) := \int_\Omega \int_\Omega w(y) \Aker(y,x) u(x) dx dy.
\end{equation}
Here the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$.
The function $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel, which exists in principle but is not easily accessible to us. In this section we describe how to extend the action of $\Aop$ to distributions, which allows us to define the impulse responses (Section \ref{sec:impulse_response_2}). Then we state the conditions on $\Aop$ that our method requires (Section \ref{sec:conditions_2}). Finally, we detail finite element discretization (Section \ref{sec:finite_element_kernel}).

@@

The action of $\Aop$ may be extended to distributions if the kernel $\Aker$ is sufficiently regular. We derive the extension as follows. Suppose $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(w) &= \int_\Omega \int_\Omega w(y) \Aker(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega w(y) \int_\Omega \Aker(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega w(y) \genericdistribution\left(\Aker(y,\cdot)\right) dy, \label{eq:action_on_distribution}
	\end{align}
\end{subequations}
where $\Aker(y,\cdot)$ denotes the function $x \mapsto \Aker(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions, and instead suppose $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ is a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in \eqref{eq:extension_to_distributions} is not valid. However, if $\Aker$ is sufficiently regular such that the function $y \mapsto \rho\left(\Aker(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the right hand side of \eqref{eq:action_on_distribution} is well-defined. Hence, we \emph{define} the action of $\Aop$ on the distribution $\rho$ to be the right hand side of \eqref{eq:action_on_distribution}. For convenience of notation, we denote this action by ``$\Aop \genericdistribution^*$,'' even if $\rho^*$ does not exist.

Let $\delta_x$ denote the delta distribution\footnote{Recall that the delta distribution $\delta_x:\mathcal{D}(\Omega)\rightarrow \mathbb{R}$ is the linear functional defined by $\delta_x(w) = w(x)$ for all $w\in \mathcal{D}(\Omega)$.} (i.e., point source, impulse) centered at the point $x \in \Omega$.
The \emph{impulse response} of $\Aop$ associated with $x$ is the function $\impulseresponse_x:\Omega \rightarrow \mathbb{R}$, 
\begin{equation}
\label{eq:impulse_response_delta_action}
\impulseresponse_x := \left( \Aop \delta_x^* \right)^*,
\end{equation}
that is formed by applying $\Aop$ to $\delta_x$, then taking the Riesz representation of the resulting linear functional. Using \eqref{eq:action_on_distribution} and the definition of the delta distribution, we see that $\phi_x$ may also be written as the function
\begin{equation}
\label{eq:impulse_response_defn}
\impulseresponse_x(y) = \Aker(y, x),
\end{equation} 
which results from keeping $x$ fixed and viewing $\Aker(y,x)$ as a function of $y$.

@@

Let $\febasis_1, \febasis_2, \dots, \febasis_\fedim$ be a set of continuous Galerkin finite element basis functions used to discretize the problem on a mesh with mesh size parameter $h$, let $V_h := \Span\left(\febasis_1, \febasis_2, \dots, \febasis_\fedim\right)$ be the corresponding finite element space under the $L^2$ inner product, and let $p_i \in \mathbb{R}^\gdim$, $i=1,\dots, \fedim$ be the Lagrange nodes associated with the functions $\febasis_i$. We assume that the finite element basis satisfies the Kronecker property, i.e., $\febasis_i(p_i)=1$ and $\febasis_i(p_j)=0$ if $i\neq j$. 
For $u_h \in V_h$ we write $\bm{u} \in \mathbb{R}^\nsamplepts_\massmatrix$ to denote the coefficient vector for $u_h$ with respect to the finite element basis, i.e.,
\begin{equation}
\label{eq:fem_coeff_basis}
u_h(x) = \sum_{i=1}^\fedim \bm{u}_i \febasis_i(x).
\end{equation}
Linear functionals $\rho_h \in V_h'$ have coefficient dual vectors $\boldsymbol{\rho}\in \mathbb{R}^\nsamplepts_{\massmatrix^{-1}}$, with entries 
\begin{equation*}
\boldsymbol{\rho}_i = \rho_h(\febasis_i), \qquad i=1,\dots,\nsamplepts. 
\end{equation*}
Here $\massmatrix \in \mathbb{R}^{\fedim \times \fedim}$ denotes the sparse finite element mass matrix which has entries
\begin{equation*}
\massmatrix_{ij}=\int_\Omega \febasis_i(x) \febasis_j(x) dx, \qquad i,j=1,\dots,\fedim,
\end{equation*}
the space $\mathbb{R}^\fedim_\massmatrix$ is $\mathbb{R}^\fedim$ with the inner product $(\mathbf{u},\mathbf{w})_\massmatrix := \mathbf{u}^T \massmatrix \mathbf{w}$, and $\mathbb{R}^\fedim_{\massmatrix^{-1}}$ is the analogous space with $\massmatrix^{-1}$ replacing $\massmatrix$. Direct calculation shows that $\mathbb{R}^\fedim_\massmatrix$ and $\mathbb{R}^\fedim_{\massmatrix^{-1}}$ are isomorphic to $V_h$ and $V_h'$ as Hilbert spaces, respectively.

After discretization, the operator $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ is replaced by an operator $A_h:V_h \rightarrow V_h'$, which becomes an operator
\begin{equation*}
\mathbf{A}:\mathbb{R}^\fedim_\massmatrix \rightarrow \mathbb{R}^\fedim_{\massmatrix^{-1}}
\end{equation*}
under the isomorphism discussed above. Our method is agnostic to the computational procedure for approximating $\Aop$ with $\mathbf{A}$. What is important is that we do not have direct access to matrix entries $\mathbf{A}_{ij}$. Rather, we have a computational procedure that allows us to compute matrix-vector products $\mathbf{u}\mapsto \mathbf{A}\mathbf{u}$ and $\mathbf{w}\mapsto \mathbf{A}^T\mathbf{w}$, and computing these matrix-vector products is costly.
Of course, matrix entries can be computed via matrix-vector products as $\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i$,
where $\mathbf{e}_j=(0,\dots,0,1,0,\dots,0)^T$ is the length $\fedim$ unit vector with one in the $j$th coordinate and zeros elsewhere. But computing the matrix-vector product $\mathbf{e}_j \mapsto \mathbf{A}\mathbf{e}_j$ is costly, and therefore wasteful if we do not use other matrix entries in the $j$th column of $\mathbf{A}$. Hence, methods for approximating $\mathbf{A}$ are computationally intractable if they require accessing scattered matrix entries from many different rows and columns of $\mathbf{A}$.

The operator $A_h:V_h \rightarrow V_h'$ can be written in integral kernel form as
\begin{equation}
(A_h u_h)(w_h) := \int_\Omega \int_\Omega w_h(y) \Aker_h(y,x) u_h(x) dx dy
\end{equation}
for some kernel $\Aker_h$ which we do not know, and which may differ from $\Aker$ due to discretization error. Since the functions in $V_h$ are continuous at $x$, the delta distribution $\delta_x$ is a continuous linear functional on $V_h$, which has a discrete dual vector $\boldsymbol{\delta}_x \in \mathbb{R}^\fedim_{\massmatrix^{-1}}$ with entries $\left(\boldsymbol{\delta}_x\right)_i = \febasis_i(x)$ for $i=1,\dots,\fedim$. Additionally, it is straightforward to verify that the Riesz representation, $\rho_h^* \in V_h$, of a functional $\rho \in V_h'$ has coefficient vector
\begin{equation*}
\boldsymbol{\rho}^* = \massmatrix^{-1} \boldsymbol{\rho}.
\end{equation*}
Therefore, the formula for the impulse response from \eqref{eq:impulse_response_delta_action} becomes 
\begin{equation}
\label{eq:discrete_kernel}
\boldsymbol{\impulseresponse}_x = \left(A_h \delta_x^*\right)^* =  \massmatrix^{-1}\mathbf{A} \massmatrix^{-1} \boldsymbol{\delta}_x,
\end{equation}
and the $(y,x)$ kernel entry of $\Aker_h$ may be written as
\begin{equation}
\label{eq:approx_kernel_entry_81}
	\Aker_h(y,x) = \boldsymbol{\delta}_y^T \boldsymbol{\impulseresponse}_x = \boldsymbol{\delta}_y^T \massmatrix^{-1}\mathbf{A} \massmatrix^{-1} \boldsymbol{\delta}_x.
\end{equation}
Now define $\mathbf{\Aker} \in \mathbb{R}^{\fedim \times \fedim}$ to be the following dense matrix of kernel entries evaluated at all pairs of Lagrange nodes:
\begin{equation}
\label{eq:Akerpcmat_entries}
\mathbf{\Aker}_{ij} := \Aker_h(p_i, p_j).
\end{equation}
Because of the Kronecker property of the finite element basis, we have $\boldsymbol{\delta}_{p_i} = \mathbf{e}_i$. Thus from \eqref{eq:approx_kernel_entry_81} we have
$
	\Aker_h(p_i,p_j) = \left(\massmatrix^{-1}\mathbf{A} \massmatrix^{-1}\right)_{ij},
$
which implies
\begin{equation}
\label{eq:boldA}
	\mathbf{A} = \massmatrix \mathbf{\Aker} \massmatrix.
\end{equation}
Broadly, our method will construct an H-matrix approximation of $\mathbf{A}$ by forming H-matrix approximations of $\boldsymbol{\Aker}$ and $\massmatrix$ (or a lumped mass version of $\massmatrix$), then multiplying these matrices with H-matrix methods per \eqref{eq:boldA}. The central challenge is that classical H-matrix construction methods require access to arbitrary matrix entries $\mathbf{\Aker}_{ij}$, but these matrix entries are not easily accessible. The bulk of our method is therefore dedicated to forming approximations of these matrix entries that can be evaluated rapidly.

@@

The direct approach to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ is to apply $\Aop$ to a point source centered at $x$ to obtain $\impulseresponse_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\impulseresponse_x$ to determine $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$. However, this direct approach is not feasible because we need to know $V(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ before we compute $\phi_x$, in order choose the point $x$. Computing $\phi_x$ in order to determine $V(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ would be extremely computationally expensive, and defeat the purpose of our algorithm, which is to reduce the computational cost by computing impulse responses in batches. Fortunately, it is possible to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ indirectly, \emph{for all points $x\in\Omega$ simultaneously}, by applying $\Aop^T$ to one constant function, $\gdim$ linear functions, and $\gdim(\gdim+1)/2$ quadratic functions (e.g., 6 total operator applies in two spatial dimensions and 10 in three spatial dimensions). This may be motivated by analogy to matrices. If $\mathbf{B}\in \mathbb{R}^{\fedim \times \fedim}$ is a matrix with $i^\text{th}$ column $\mathbf{b}_i$ and $\mathbf{f} \in \mathbb{R}^\fedim$, then
\begin{equation*}
\mathbf{B}^T \mathbf{w} = \begin{bmatrix}
\horzbar & \mathbf{b}_1^T & \horzbar \\
\horzbar & \mathbf{b}_2^T & \horzbar \\
& \vdots & \\
\horzbar & \mathbf{b}_\fedim^T & \horzbar
\end{bmatrix}
\mathbf{w} = 
\begin{bmatrix}
\mathbf{b}_1^T \mathbf{w} \\
\mathbf{b}_2^T \mathbf{w} \\
\vdots \\
\mathbf{b}_\fedim^T \mathbf{w}
\end{bmatrix}.
\end{equation*}
By computing one matrix-vector product of $\mathbf{B}^T$ with $\mathbf{w}$, we compute the inner product of each column of $\mathbf{B}$ with $\mathbf{w}$ simultaneously. The operator case is analogous, with the impulse response $\phi_x$ taking the place of a matrix column. We have
\begin{equation}
\label{eq:operator_simultaneous_ips}
\left(\Aop^T w\right)^*(x) = \int_\Omega \Aker(y,x) w(y) dy = \left(\phi_x, w\right)_{L^2(\Omega)}.
\end{equation}
By computing one operator application of $\Aop^T$ to $w$, we compute the inner product of each impulse response $\phi_x$ with $w$, for all points $x$ simultaneously.

Let $C$, $L^i$, and $Q^{ij}$ be the following constant, linear, and quadratic functions:
\begin{equation*}
C(x) := 1, \qquad
L^i(x) := x^i, \qquad
Q^{ij}(x) := x^i x^j
\end{equation*}
for $i,j=1,\dots,\gdim$. Using the definition of $\spatialvol$ in \eqref{eq:define_vol} and using \eqref{eq:operator_simultaneous_ips}, we have
\begin{equation*}
	\spatialvol(x) = \int_\Omega \phi_x(y) C(y) ~ dy = \left(\phi_x, C\right)_{L^2(\Omega)} = \left(\mathcal{A}^T C\right)^*(x).
\end{equation*}
Hence, we may compute $\spatialvol(x)$ for all $x$ simultaneously by applying $\Aop^T$ to $C$. Analogous manipulations show that $\spatialmean(x)$ and $\spatialcov(x)$ may be computed for all points $x$ simultaneously by applying $\Aop^T$ to the functions $L^i$ and $Q^{ij}$, respectively. To summarize the results of doing this, we have
\begin{subequations}
	\label{eq:vol_mean_var}
	\begin{align}
	\spatialvol =& \left(\Aop^T C\right)^* \\
	\spatialmean^i =& \left(\Aop^T L^i\right)^* / \spatialvol \\
	\spatialcov^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \spatialvol - \spatialmean^i\cdot \spatialmean^j
	\end{align}
\end{subequations}
for $i,j=1,\dots, \gdim$. Here $u/w$ denotes pointwise division of functions, $\left(u/w\right)(x) = u(x)/w(x)$, and $u\cdot w$ denotes pointwise multiplication of functions, $(u \cdot w)(x) = u(x)w(x)$.
%The resulting method for computing discretized versions of $\spatialvol$, $\spatialmean$, and $\spatialcov$ is shown in Algorithm \ref{alg:varhpi_mean_cov}.
